### 2025-02-12 misc/expo/entropy.tex
Updates for 2025

\def\eps{\epsilon}
\def\FF{{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\Bohr{\op{\rm Bohr}}
\def\Eta{\op{\bf H}}
\def\Etaseven{\op{\sevenbf H}}
\def\II{\op{\bf I}}
\def\dd{\op{\bf d}}
\def\ee{\op{\bf e}}
\def\P{{\cal P}}
\def\given{\mathbin{|}}
\def\argmax{\limitop{\rm arg$\,$max}}
\def\dTV{d_{\rm TV}}
\def\advthm{\the\sectcount.\the\thmcount\global\advance \thmcount by 1}
\def\advsect{\global\advance\sectcount by 1\section\the\sectcount\global\thmcount=1. }
\sectcount=0

@@

We are now done if we can show that entropy is nonnegative.
This is a corollary of the following lemma. The following proof is a modification of
one due to S.~Eberhard.

@@

\proof The hypotheses give the three inequalities
$$\Eta\{X\} \le \Eta\{Z,W\}, \quad \Eta\{Y\}\le \Eta\{Z\},\quad\hbox{and}\quad \Eta\{Y\}\le \Eta\{W\}.$$
From this we see that
$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z,W\} + \Eta\{Z\} + \Eta\{Y\}.$$
But then since conditioning does not increase entropy, we have
$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z\} + 2\Eta\{Y\},$$
whence dividing both sides by $2$ completes the proof.\slug

@@

We now define an entropic version of $E(A,B)$. Let $X$ and $Y$ be discrete random variables taking values
in the same
abelian group. Let $(X_1, Y_1)$ and $(X_2, Y_2)$ be conditionally independent trials of $(X,Y)$ relative to
$X+Y$. These  $X_1 + Y_1 = X_2 + Y_2 = A+B$.
The {\it entropic additive energy} between $X$ and $Y$ is
$$ \ee\{X,Y\} = \Eta\{X_1, Y_1, X_2, Y_2\}.$$
This definition makes clear the analogy between this value and the additive energy of sets, but
by conditional independence and the fact that $(X_1, Y_1, X_2, Y_2)$ determines $X_1+Y_1 = X+Y$,
we can rewrite
$$\eqalign{
\ee\{X,Y\} &= \Eta\{X_1, Y_1, X_2, Y_2, X+Y\} \cr
&= 2\Eta\{X,Y,X+Y\} - \Eta\{X+Y\} \cr
&= 2\Eta\{X,Y\} - \Eta\{X+Y\},\cr
}$$
where in the second equality we applied~\refeq{\eqcondindep}. This formula is something we will use
often, as it makes no direct mention of the variables $(X_1, Y_1)$ and $(X_2,Y_2)$.

@@

\edef\propnaive{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $X$ and $Y$ be discrete random variables taking values in the
same abelian group. If
\global\edef\eqenergybound{\the\eqcount}
$$\ee\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2}\Eta\{Y\} - \log K,\adveq$$
for some constant $K$, then
\global\edef\eqsumbound{\the\eqcount}
$$\Eta\{ X+Y\} \le {1\over 2} \Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K.\adveq$$
If one adds the further assumption that
$\Eta\{X,Y\} \ge \Eta\{X\} + \Eta\{Y\} - C$,
then~{\rm \refeq{\eqsumbound}} implies~{\rm \refeq{\eqenergybound}} with a worse constant, namely,
we may only conclude
\global\edef\eqworsebound{\the\eqcount}
$$\ee\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2}\Eta\{Y\} - \log K - 2C.\adveq$$
In particular, if $X$ and $Y$ are independent, then we can recover~{\rm \refeq{\eqenergybound}}
from~{\rm \refeq{\eqsumbound}}.

\proof Assuming the lower bound on the additive energy, we have
$$2\Eta\{X,Y\} - \Eta\{X+Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} - \log K,$$
so
$$\eqalign{
\Eta\{X+Y\} &\le 2\Eta\{X,Y\} - {3\over 2} \Eta\{X\} - {3\over 2} \Eta\{Y\} + \log K \cr
&\le 2\Eta\{X\} + 2\Eta\{Y\} - {3\over 2} \Eta\{X\} - {3\over 2} \Eta\{Y\} + \log K \cr
&= {1\over 2} \Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K .\cr
}$$
On the other hand, assuming this upper bound on $\Eta\{X+Y\}$, we have
$$\eqalign{
\ee\{X,Y\} &= 2\Eta\{X,Y\} - \Eta\{X+Y\}  \cr
&\ge 2\Eta\{X,Y\} - {1\over 2} \Eta\{X\} - {1\over 2} \Eta\{Y\} - \log K ,\cr
}$$
and if $2\Eta\{X,Y\} \ge 2 \Eta\{X\} + 2\Eta\{Y\} - 2C$, then~\refeq{\eqworsebound} follows
directly.\slug

@@

Simply put, in the entropic setting the bound~\refeq{\eqenergybound} is
stronger than~\refeq{\eqsumbound} because the entropy $\Eta\{X,Y\}$ of the joint distribution appears in
the formula for $\ee\{X,Y\}$, whereas~\refeq{\eqsumbound} says nothing whatsoever about this joint distribution.

@@

\edef\lematwo{\the\sectcount.\the\thmcount}
\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma A.2).
Let $X$ and $Y$ be discrete random variables taking values in the same abelian
group. Let $(X_1, Y_1)$ and $(X_2,Y_2)$ be conditionally
independent trials of $(X,Y)$ relative to $X+Y$. Then we have
$$\max\bigl(\Eta\{X_1- X_2\}, \Eta\{X_1-Y_2\}\bigr) \le \Eta\{X+Y\} + 2\II\{X:Y\}.$$
The right-hand side of this expression can also be written $2\Eta\{X\} + 2\Eta\{Y\} - \ee\{X,Y\}$.

@@

\parenproclaim Theorem {\advthm} (Entropic Balog--Szemer\'edi--Gowers theorem).
Let $X$ and $Y$ be discrete random variables taking values in the same abelian group, and
suppose that
$$\ee\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} + \log K$$
for some constant $K$.
Then letting $(X_1, Y_1)$ and $(X_2,Y_2)$ be conditionally
independent trials of $(X,Y)$ relative to $X+Y$, we have
$$ \Eta\{ X_1 \given X+Y\} \ge \Eta\{X\} - 2\log K$$
and
$$ \Eta\{ Y_2 \given X+Y\} \ge \Eta\{Y\} - 2\log K.$$
Furthermore, the variables $X_1$ and $Y_2$ are conditionally independent relative to $X+Y$, and
we have
$$\Eta\{ X_1 + Y_2 \given X+Y\} \le {1\over 2}\Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K.$$

\proof Using the coupling $X+Y = X_1+Y_1 = X_2 + Y_2$, we have
$$\eqalign{
\Eta\{X_1\given X + Y\} &= \Eta\{X_1, X_1+Y_1\} - \Eta\{X+Y\} \cr
&= \Eta\{X,Y\} - \Eta\{X+Y\} \cr
&= \ee\{X,Y\} - \Eta\{X,Y\} \cr
&\ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} - \log K - \Eta\{X,Y\} \cr
&\ge {1\over 2} \Eta\{X\} + {1\over 2}\Eta\{Y\} -\log K.\cr
}$$
where in the fourth line we used the hypothesis on $\ee\{X,Y\}$, and in the last line we observed
that $\Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \ge 0$. The bound
$$\Eta\{Y_2, X+Y\} \ge {1\over 2} \Eta\{X\} + {1\over 2}\Eta\{Y\} -\log K$$
is shown in the exact same way; only the first step differs.

@@

It remains to prove the upper bound on $\Eta\{X_1 + Y_2 \given X+Y\}$.
Note that $(X_1, Y_2, X+Y)$ and $(X_1 - X_2, X+Y)$ jointly determine $(X_1, X_2, X+Y)$.
Then given $X_1 - X_2$ and $X+Y$ we can calculate
$$X_1 + Y_2 = X_1 - X_2 + X_2 + Y_2 = (X_1 - X_2) + (X+Y),$$
so $(X_1, Y_2, X+Y)$ and $(X_1 - X_2, X+Y)$
each separately determine $(X_1 + Y_2, X+Y)$. Hence the submodularity inequality yields
$$ \Eta\{X_1, X_2, X+Y\} + \Eta\{X_1+Y_2, X+Y\} \le \Eta\{X_1, Y_2, X+Y\} + \Eta\{X_1- X_2, X+Y\}.$$
From $(X_1, X_2, X+Y)$
we can calculate $Y_1 = X+Y-X_1$ and $Y_2 = X+Y-X_2$, so this triple and the triple
$(X_1, X_2, Y_1, Y_2)$ determine each other. So the first term above is simply the additive energy
between $X$ and $Y$; that is
$$ \Eta\{X_1, X_2, X+Y\} = \Eta\{X_1, X_2, Y_1, Y_2\} = 2\Eta\{X,Y\} - \Eta\{X+Y\}.$$
Now since $X_1$ and $Y_2$ are conditionally independent relative to $X+Y$, we have
$$\eqalign{
\Eta\{X_1, Y_2, X+Y\} &= \Eta\{X_1, X+Y\} + \Eta\{Y_2, X+Y\} - \Eta\{X+Y\} \cr
&= \Eta\{X, X+Y\} + \Eta\{Y, X+Y\} - \Eta\{X+Y\} \cr
&= 2\Eta\{X,Y\} - \Eta\{X+Y\} \cr
}$$
For the last term above we split
$$\Eta\{X_1 - X_2, X+Y\} = \Eta\{X_1 - X_2 \given X+Y\} - \Eta\{X+Y\}.$$
Putting everything together, we obtain
$$\eqalign{
2\Eta\{X,Y\} - \Eta\{X+&Y\} + \Eta\{X_1 + Y_2, X+Y\} \cr
&\le 2\Eta\{X,Y\} + \Eta\{X_1-X_2\given X+Y\} - 2\Eta\{X+Y\},\cr
}$$
so that
$$\Eta\{ X_1 + Y_2\given X+Y\} \le \Eta\{X_1 - X_2\given X+Y\} - 2\Eta\{X+Y\}\le \Eta\{X_1 - X_2\}.$$
The previous lemma then gives
$$\Eta\{ X_1 + Y_2\given X+Y\} \le 2\Eta\{X\} + 2\Eta\{Y\} - \ee\{X,Y\},$$
and from our lower bound on $\ee\{X,Y\}$, we conclude that
$$\Eta\{ X_1 + Y_2\given X+Y\} \le {1\over 2} \Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K,$$
which is what we wanted to show.\slug

@@

>>>>

\def\eps{\epsilon}
\def\FF{\hbox{\bf F}}
\def\bar#1{\overline{#1}}
\def\hat#1{\widehat{#1}}
\def\norm#1{|\!|#1|\!|}
\def\bignorm#1{\big|\!\big|#1\big|\!\big|}
\def\Norm#1{\Big|\!\Big|#1\Big|\!\Big|}
\def\normm#1{\bigg|\!\bigg|#1\bigg|\!\bigg|}
\def\Bohr{\op{\rm Bohr}}
\def\Eta{\op{\bf H}}
\def\Etaseven{\op{\sevenbf H}}
\def\II{\op{\bf I}}
\def\dd{\op{\bf d}}
\def\AA{\op{\bf A}}
\def\P{{\cal P}}
\def\given{\mathbin{|}}
\def\argmax{\limitop{\rm arg$\,$max}}
\def\dTV{d_{\rm TV}}
\def\advthm{\the\sectcount.\the\thmcount\global\advance \thmcount by 1}
\def\advsect{\global\advance\sectcount by 1\section\the\sectcount\global\thmcount=1. }
\sectcount=0

@@

We are now done if we can show that entropy is nonnegative.
This is a corollary of the following lemma, whose proof is a modification of
one due to S.~Eberhard.

@@

\proof The hypotheses give the three inequalities
$$\Eta\{X\} \le \Eta\{Z,W\}, \quad \Eta\{Y\}\le \Eta\{Z\},\quad\hbox{and}\quad \Eta\{Y\}\le \Eta\{W\}.$$
From this we see that
$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z,W\} + \Eta\{Z\} + \Eta\{W\}.$$
But then since conditioning does not increase entropy, we have
$$2\Eta\{X\} + 2\Eta\{Y\} \le 2\Eta\{Z\} + 2\Eta\{W\},$$
whence dividing both sides by $2$ completes the proof.\slug

@@

We now define an entropic version of $E(A,B)$. Let $X$ and $Y$ be discrete random variables taking values
in the same
abelian group. Let $(X_1, Y_1)$ and $(X_2, Y_2)$ be conditionally independent trials of $(X,Y)$ relative to
$X+Y$. These  $X_1 + Y_1 = X_2 + Y_2 = A+B$.
The {\it entropic additive energy} between $X$ and $Y$ is
$$ \AA\{X,Y\} = \Eta\{X_1, Y_1, X_2, Y_2\}.$$
This definition makes clear the analogy between this value and the additive energy of sets, but
by conditional independence and the fact that $(X_1, Y_1, X_2, Y_2)$ determines $X_1+Y_1 = X+Y$,
we can rewrite
$$\eqalign{
\AA\{X,Y\} &= \Eta\{X_1, Y_1, X_2, Y_2, X+Y\} \cr
&= 2\Eta\{X,Y,X+Y\} - \Eta\{X+Y\} \cr
&= 2\Eta\{X,Y\} - \Eta\{X+Y\},\cr
}$$
where in the second equality we applied~\refeq{\eqcondindep}. This formula is something we will use
often, as it makes no direct mention of the variables $(X_1, Y_1)$ and $(X_2,Y_2)$.

@@

\edef\propnaive{\the\sectcount.\the\thmcount}
\proclaim Proposition \advthm. Let $X$ and $Y$ be discrete random variables taking values in the
same abelian group. If
\global\edef\eqenergybound{\the\eqcount}
$$\AA\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2}\Eta\{Y\} - \log K,\adveq$$
for some constant $K$, then
\global\edef\eqsumbound{\the\eqcount}
$$\Eta\{ X+Y\} \le {1\over 2} \Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K.\adveq$$
If one adds the further assumption that
$\Eta\{X,Y\} \ge \Eta\{X\} + \Eta\{Y\} - C$,
then~{\rm \refeq{\eqsumbound}} implies~{\rm \refeq{\eqenergybound}} with a worse constant, namely,
we may only conclude
\global\edef\eqworsebound{\the\eqcount}
$$\AA\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2}\Eta\{Y\} - \log K - 2C.\adveq$$
In particular, if $X$ and $Y$ are independent, then we can recover~{\rm \refeq{\eqenergybound}}
from~{\rm \refeq{\eqsumbound}}.

\proof Assuming the lower bound on the additive energy, we have
$$2\Eta\{X,Y\} - \Eta\{X+Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} - \log K,$$
so
$$\eqalign{
\Eta\{X+Y\} &\le 2\Eta\{X,Y\} - {3\over 2} \Eta\{X\} - {3\over 2} \Eta\{Y\} + \log K \cr
&\le 2\Eta\{X\} + 2\Eta\{Y\} - {3\over 2} \Eta\{X\} - {3\over 2} \Eta\{Y\} + \log K \cr
&= {1\over 2} \Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K .\cr
}$$
On the other hand, assuming this upper bound on $\Eta\{X+Y\}$, we have
$$\eqalign{
\AA\{X,Y\} &= 2\Eta\{X,Y\} - \Eta\{X+Y\}  \cr
&\ge 2\Eta\{X,Y\} - {1\over 2} \Eta\{X\} - {1\over 2} \Eta\{Y\} - \log K ,\cr
}$$
and if $2\Eta\{X,Y\} \ge 2 \Eta\{X\} + 2\Eta\{Y\} - 2C$, then~\refeq{\eqworsebound} follows
directly.\slug

@@

Simply put, in the entropic setting the bound~\refeq{\eqenergybound} is
stronger than~\refeq{\eqsumbound} because the entropy $\Eta\{X,Y\}$ of the joint distribution appears in
the formula for $\AA\{X,Y\}$, whereas~\refeq{\eqsumbound} says nothing whatsoever about this joint distribution.

@@

\edef\lematwo{\the\sectcount.\the\thmcount}
\parenproclaim Lemma {\advthm} ({\rm\ref{pfr2023},} Lemma A.2).
Let $X$ and $Y$ be discrete random variables taking values in the same abelian
group. Let $(X_1, Y_1)$ and $(X_2,Y_2)$ be conditionally
independent trials of $(X,Y)$ relative to $X+Y$. Then we have
$$\max\bigl(\Eta\{X_1- X_2\}, \Eta\{X_1-Y_2\}\bigr) \le \Eta\{X+Y\} + 2\II\{X:Y\}.$$
The right-hand side of this expression can also be written $2\Eta\{X\} + 2\Eta\{Y\} - \AA\{X,Y\}$.

@@

\parenproclaim Theorem {\advthm} (Entropic Balog--Szemer\'edi--Gowers theorem).
Let $X$ and $Y$ be discrete random variables taking values in the same abelian group, and
suppose that
$$\AA\{X,Y\} \ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} + \log K$$
for some constant $K$.
Then letting $(X_1, Y_1)$ and $(X_2,Y_2)$ be conditionally
independent trials of $(X,Y)$ relative to $X+Y$, we have
$$ \Eta\{ X_1 \given X+Y\} \ge \Eta\{X\} - 2\log K$$
and
$$ \Eta\{ Y_2 \given X+Y\} \ge \Eta\{Y\} - 2\log K.$$
Furthermore, the variables $X_1$ and $Y_2$ are conditionally independent relative to $X+Y$, and
we have
$$\Eta\{ X_1 + Y_2 \given X+Y\} \le {1\over 2}\Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K.$$

\proof Using the coupling $X+Y = X_1+Y_1 = X_2 + Y_2$, we have
$$\eqalign{
\Eta\{X_1\given X + Y\} &= \Eta\{X_1, X_1+Y_1\} - \Eta\{X+Y\} \cr
&= \Eta\{X,Y\} - \Eta\{X+Y\} \cr
&= \AA\{X,Y\} - \Eta\{X,Y\} \cr
&\ge {3\over 2} \Eta\{X\} + {3\over 2} \Eta\{Y\} - \log K - \Eta\{X,Y\} \cr
&\ge {1\over 2} \Eta\{X\} + {1\over 2}\Eta\{Y\} -\log K.\cr
}$$
where in the fourth line we used the hypothesis on $\AA\{X,Y\}$, and in the last line we observed
that $\Eta\{X\} + \Eta\{Y\} - \Eta\{X,Y\} \ge 0$. The bound
$$\Eta\{Y_2, X+Y\} \ge {1\over 2} \Eta\{X\} + {1\over 2}\Eta\{Y\} -\log K$$
is shown in the exact same way; only the first step differs.

@@

It remains to prove the upper bound on $\Eta\{X_1 + Y_2 \given X+Y\}$.
Note that $(X_1, Y_2, X+Y)$ and $(X_1 - X_2, X+Y)$ jointly determine $(X_1, X_2, X+Y)$.
Then given $X_1 - X_2$ and $X+Y$ we can calculate
$$X_1 + Y_2 = X_1 - X_2 + X_2 + Y_2 = (X_1 - X_2) + (X+Y),$$
so $(X_1, Y_2, X+Y)$ and $(X_1 - X_2, X+Y)$
each separately determine $(X_1 + Y_2, X+Y)$. Hence the submodularity inequality yields
$$ \Eta\{X_1, X_2, X+Y\} + \Eta\{X_1+Y_2, X+Y\} \le \Eta\{X_1, Y_2, X+Y\} + \Eta\{X_1- X_2, X+Y\}.$$
From $(X_1, X_2, X+Y)$
we can calculate $Y_1 = X+Y-X_1$ and $Y_2 = X+Y-X_2$, so this triple and the triple
$(X_1, X_2, Y_1, Y_2)$ determine each other. So the first term above is simply the additive energy
between $X$ and $Y$; that is
$$ \Eta\{X_1, X_2, X+Y\} = \Eta\{X_1, X_2, Y_1, Y_2\} = 2\Eta\{X,Y\} - \Eta\{X+Y\}.$$
Now since $X_1$ and $Y_2$ are conditionally independent relative to $X+Y$, we have
$$\eqalign{
\Eta\{X_1, Y_2, X+Y\} &= \Eta\{X_1, X+Y\} + \Eta\{Y_2, X+Y\} - \Eta\{X+Y\} \cr
&= \Eta\{X, X+Y\} + \Eta\{Y, X+Y\} - \Eta\{X+Y\} \cr
&= 2\Eta\{X,Y\} - \Eta\{X+Y\} \cr
}$$
For the last term above we split
$$\Eta\{X_1 - X_2, X+Y\} = \Eta\{X_1 - X_2 \given X+Y\} - \Eta\{X+Y\}.$$
Putting everything together, we obtain
$$\eqalign{
2\Eta\{X,Y\} - \Eta\{X+&Y\} + \Eta\{X_1 + Y_2, X+Y\} \cr
&\le 2\Eta\{X,Y\} + \Eta\{X_1-X_2\given X+Y\} - 2\Eta\{X+Y\},\cr
}$$
so that
$$\Eta\{ X_1 + Y_2\given X+Y\} \le \Eta\{X_1 - X_2\given X+Y\} - 2\Eta\{X+Y\}\le \Eta\{X_1 - X_2\}.$$
The previous lemma then gives
$$\Eta\{ X_1 + Y_2\given X+Y\} \le 2\Eta\{X\} + 2\Eta\{Y\} - \AA\{X,Y\},$$
and from our lower bound on $\AA\{X,Y\}$, we conclude that
$$\Eta\{ X_1 + Y_2\given X+Y\} \le {1\over 2} \Eta\{X\} + {1\over 2} \Eta\{Y\} + \log K,$$
which is what we wanted to show.\slug

@@

\advsect Sum-product phenomena

We prove the following lemma.

\proclaim Lemma \advthm.
Let $X$ and $Y$ be random variables taking values in $\FF_p^*$ for some prime $p$.
There exists $u\in \FF_p^*$ such that
$$\Eta\{X,uY\} \ge {\Eta\{X,Y\}\over 2^{\Etaseven\{X,Y\}}} - 1,$$

\proof Let $U$ be a random variable uniform on $\FF_p^*$ and independent of both $X$ and $Y$.
Independence gives us the identity $\Eta\{X+UY\} = \Eta\{U\} = \log(p-1)$.
We let $(X_1, Y_1)$, $(X_2,Y_2)$, and $U$ be conditionally independent trials
of $(X,Y)$, $(X,Y)$, and $U$ respectively, relative to $X+UY$.
Put
$$H = \Eta\{X_1, Y_1, X_2, Y_2, U\}.$$
Let $W = 1$ if $X_1 = X_2$ and $W=0$ otherwise. Then since $X_1 + UY_1 = X+UY$, we have
$$\eqalign{
H &= \Eta\{X_1, Y_1, X_2, Y_2, W, X+UY\} \cr
&= \Eta\{U\} + \Eta\{W\} + \Eta\{X_1, Y_1, X_2, Y_2 \given W, X+UY\}.\cr
}$$
If $X_1 = X_2$, then $Y_1 = Y_2$, so by conditional independence,
$$\eqalign{
H &= \Eta\{U\} + \Eta\{W\} +  \pr\{X_1 = X_2\given X+UY\} \Eta\{X_1, Y_1 \given X+UY\} \cr
&\qquad+ \pr\{X_1 \ne X_2 \given X+UY\} \Eta\{X_1,Y_1, X_2,Y_2\given X+UY\}.\cr
&= \Eta\{U\} + \Eta\{W\} + \pr\{X_1 = X_2\}\bigl( \Eta\{X,Y\given X+UY\}\bigr) \cr
&\qquad
+\pr\{X_1 \ne X_2\given X+UY\}\bigl( 2\Eta\{X,Y\given X+UY\} \bigr) \cr
&= \Eta\{U\} + \Eta\{W\} + \Eta\{X,Y\given X+UY\}\cr
&\qquad+\pr\{X_1 \ne X_2\given X+UY\}\Eta\{X,Y \given X+UY\} \cr
}$$
But
$$\Eta\{X,Y\given X+UY\} = \Eta\{X,Y, X+UY\} - \Eta\{X+UY\} = \Eta\{X,Y\}$$
by independence of $(X,Y)$ and $U$, so
$$H =\bigl(1+\pr\{X_1\ne X_2\given X+UY\}\}\bigr) \Eta\{X,Y\} + \Eta\{U\} + \Eta\{W\}.$$

On the other hand, by invertibility of each $u\in \FF_p^*$, we have
$$\eqalign{
H - \Eta\{U\} &= \Eta\{X_1, Y_1, X_2, Y_2\given U\} \cr
&= {1\over p-1} \sum_{u\in \FF_p^*} \Eta\{X_1, Y_1, X_2, Y_2\given U=u\} \cr
&= {1\over p-1} \sum_{u\in \FF_p^*} \Eta\{X_1, uY_1, X_2, uY_2\} \cr
&= {1\over p-1} \sum_{u\in \FF_p^*} \AA\{X, uY\}.\cr
}$$
Consequently there exists $u\in \FF_p^*$ such that
$$\eqalign{
\AA\{X,uY\} &\le H-\Eta\{U\} \cr
&= \bigl(1+\pr\{X_1\ne X_2\given X+UY\}\bigr) \Eta\{X,Y\} + \Eta\{W\} \cr
&\le \bigl(1+\pr\{X_1\ne X_2\given X+UY\}\bigr) \Eta\{X,Y\} + 1, \cr
}$$
and for this choice of $u$ we have
$$\Eta\{X+uY\} \ge \pr\{X_1 = X_2\given X+UY\} \Eta\{X,Y\} - 1.$$
But
$$\pr\{X_1 = X_2\given X+UY\} = \pr\bigl\{ (X_1,Y_1) = (X_2, Y_2) \given X+UY\bigr\},$$
and by [TODO: add lemma], we have
$$\eqalign{
\pr\{ (X_1,Y_1) = (X_2, Y_2) \given X+UY\} &\ge 2^{-\Etaseven\{X, Y \given X+UY\}} \cr
&= 2^{-\Etaseven\{X, Y\}},\cr
}$$
which immediately yields the desired inequality.\slug
